---
title: "Guided Project: Answering Business Questions using SQL"
author: "Chun Hoe Leung"
output: html_notebook
---

### 1. Intro
I'll work with a modified version of a database called Chinook. The Chinook database contains information about various elements in a fictional digital music shop, such as artists, songs, and albums, as well as the shop's employees, customers, and the customers' purchases. All this information is contained in 11 tables. The Chinook database is provided in the chinook.db file. 
![Alt text](C:/Users/Familie Leung/Documents/CHL/DataQuest R/Guided Project Answering Business Questions using SQL/chinook-schema.svg)


### 2. Creating Helper Functions
**Import the RSQLite and DBI libraries.**
```{r}
library(RSQLite)
library(DBI)

db<-'C:/Users/Familie Leung/Documents/CHL/DataQuest R/Guided Project Answering Business Questions using SQL/chinook.db'
```
**Create a run_query() function that takes in a SQL query as an argument and returns the result in a dataframe.**
```{r}
run_query<-function(query) {
  conn=dbConnect(SQLite(),db)
  result<-dbGetQuery(conn,query)
  dbDisconnect(conn)
  return(result)
}
```

**Create a show_tables() function that calls the run_query() function to return a list of all tables and views in the database.**
```{r}
show_tables<-function(){
  q="SELECT name,type FROM sqlite_master WHERE type IN ('table','view')"
  return(run_query(q))
}
```
**Run the show_tables() function.**
```{r}
show_tables()
```

### 3. Selecting Albums to Purchase
**Write a query that returns each genre, with the number of tracks sold in the USA:in absolute numbers and in percentages**
```{r}
query1<-'
WITH
  tracks_sold_usa AS
  (
  SELECT 
    COUNT(il.track_id) tracks_sold_usa_total
  FROM invoice_line il
  INNER JOIN invoice i ON il.invoice_id=i.invoice_id
  GROUP BY i.billing_country
  HAVING i.billing_country="USA"
  )
  
SELECT 
  g.name genre,
  COUNT(il.track_id) tracks_sold_numbers,
  (CAST(COUNT(il.track_id)as Float)/(SELECT * FROM tracks_sold_usa))*100 tracks_sold_percentages,
  i.billing_country
FROM invoice_line il
INNER JOIN track t ON il.track_id=t.track_id
INNER JOIN genre g ON t.genre_id=g.genre_id
INNER JOIN (
            SELECT * FROM invoice
            WHERE billing_country="USA"
           ) i ON il.invoice_id=i.invoice_id
GROUP BY g.genre_id
ORDER BY 3 DESC
'
run_query(query1)
```
**Create a plot to show this data**
```{r}
library(ggplot2)
ggplot(data=run_query(query1))+aes(x=genre,y=tracks_sold_percentages,fill=genre)+geom_bar(stat='identity')+coord_flip()
```

**Write a paragraph that interprets the data and makes a recommendation for the three artists whose albums we should purchase for the store, based on sales of tracks from their genres.**

Among the genres represented in our list of 4 albums, punk, blues and pop are the highest rated. Therefore, we should recommend:

- Red Tone (Punk)
- Slim Jim Bites (Blues)
- Meteor and the Girls (Pop)

By far though, rock makes up the majority of the sales. To better capture sales in the USA, we might want to ask the record label if they have any up-and-coming rock bands.

### 4. Analyzing Employee Sales Performance

**Write a query that finds the total dollar amount of sales assigned to each sales support agent within the company. Add any extra attributes for that employee that you find are relevant to the analysis.**
```{r}
query2<-'
SELECT DISTINCT title FROM employee
'
run_query(query2)
```

```{r}
query3<-'
WITH 
  sales_support_agent AS
    (
    SELECT * FROM employee
    WHERE title="Sales Support Agent"
    ),
  total_sales_per_customer AS
    (
    SELECT 
      c.*,
      COUNT(i.invoice_id) number_of_invoices,
      SUM(i.total) total
    FROM customer c
    INNER JOIN invoice i ON c.customer_id=i.customer_id
    GROUP BY c.customer_id
    )
SELECT
  ssa.employee_id,
  ssa.first_name||" "||ssa.last_name employee_name,
  ssa.title,
  ssa.birthdate,
  ssa.hire_date,
  COUNT(tspc.customer_id) number_of_customers,
  COUNT(DISTINCT(tspc.country)) number_of_countries,
  SUM(tspc.number_of_invoices) number_of_invoices,
  SUM(tspc.total) total_sales
FROM sales_support_agent ssa
LEFT JOIN total_sales_per_customer tspc ON ssa.employee_id=tspc.support_rep_id
GROUP BY ssa.employee_id
'
run_query(query3)
```

**Create a plot of the results of your query.**
```{r}
ggplot(data=run_query(query3))+aes(x=employee_name,y=total_sales,fill=employee_name)+geom_bar(stat='identity')
```

**Write a short statement describing your results, and provide a possible interpretation.**
The company Chinook counts 3 sales support agents:
-Jane Peacock 
-Margaret Park
-Steve Johnson
Looking from [A] total sales, [B] number of customers, [C] number of invoices perspectives, Jane Peacock (also the youngest) outperforms from the rest of her collegues. Followed by Margaret Park (oldest) and Steve Johnson.
However there might be more reasons behind these figures that we haven't analyzed, such as the relationship between their language skills and their customers. Steve serves customers from 13 different countries (most), while Jane serves only 10 (least).

### 5. Analyzing Sales by Country

**Write a query that collates data on purchases from different countries. For each country, include: [1] total number of customers,[2] total value of sales,[3] average value of sales per customer, [4] average order value. Where a country has only one customer, collect them into an "Other" group.**
```{r}
query4<-'
WITH
  country_one_customer AS
    (
    SELECT 
      C.country,
      CASE
        WHEN COUNT(DISTINCT(c.customer_id))=1 THEN "Other"
        ELSE c.country
        END
        AS other_country2
      FROM customer c
      GROUP BY country
      HAVING other_country2="Other"
    )

SELECT
  CASE
    WHEN c.country IN (SELECT country FROM country_one_customer) THEN "Other"
    ELSE c.country
    END
    AS country2,
  COUNT(DISTINCT(c.customer_id)) total_number_of_customers,
  COUNT(c.customer_id) total_number_of_invoices,
  SUM(i.total) total_value_of_sales,
  SUM(i.total)/COUNT(DISTINCT(c.customer_id)) average_value_of_sales_per_customer,
  SUM(i.total)/COUNT(c.customer_id) average_value_of_sales_per_invoice
FROM customer c
LEFT JOIN invoice i ON c.customer_id=i.customer_id
GROUP BY country2
ORDER BY 4 DESC
'
run_query(query4)
```
**The results should be sorted by the total sales from highest to lowest, with the "Other" group at the very bottom.**
```{r}
query5<-'
WITH
  country_one_customer AS
    (
    SELECT 
      C.country,
      CASE
        WHEN COUNT(DISTINCT(c.customer_id))=1 THEN "Other"
        ELSE c.country
        END
        AS other_country2
      FROM customer c
      GROUP BY country
      HAVING other_country2="Other"
    )

SELECT
  CASE
    WHEN c.country IN (SELECT country FROM country_one_customer) THEN "Other"
    ELSE c.country
    END
    AS country2,
  COUNT(DISTINCT(c.customer_id)) total_number_of_customers,
  COUNT(c.customer_id) total_number_of_invoices,
  SUM(i.total) total_value_of_sales,
  SUM(i.total)/COUNT(DISTINCT(c.customer_id)) average_value_of_sales_per_customer,
  SUM(i.total)/COUNT(c.customer_id) average_value_of_sales_per_invoice,
  CASE
    WHEN c.country IN (SELECT country FROM country_one_customer) THEN 1
    ELSE 0
    END
    AS sort
FROM customer c
LEFT JOIN invoice i ON c.customer_id=i.customer_id
GROUP BY country2
ORDER BY 7 ASC, 4 DESC
'
run_query(query5)
```

### 6. Visualizing Sales by Country

**For each dimension, create a visualization which demonstrates the data we collated in the previous step**
```{r}
ggplot(data=run_query(query5))+aes(x = reorder(country2,total_value_of_sales),y=total_value_of_sales,fill=country2)+geom_bar(stat='identity')+coord_flip()+labs(title="Total Sales by country",x = "Country",y = "Total Sales")

```
```{r}
ggplot(data=run_query(query5))+aes(x = reorder(country2,total_number_of_customers),y=total_number_of_customers,fill=country2)+geom_bar(stat='identity')+coord_flip()+labs(title="Total Customers by country",x = "Country",y = "Total Sales")
```



```{r}
ggplot(data=run_query(query5))+aes(x = reorder(country2,average_value_of_sales_per_customer),y=average_value_of_sales_per_customer,fill=country2)+geom_bar(stat='identity')+coord_flip()+labs(title="Average Sales per Customer by country",x = "Country",y = "Total Sales")
```
```{r}
ggplot(data=run_query(query5))+aes(x = reorder(country2,total_number_of_customers),y=total_number_of_customers,fill=country2)+geom_bar(stat='identity')+coord_flip()+labs(title="Total customers by country",x = "Country",y = "Total Sales")
```

**Write a few sentences interpreting your data and visualizations, and make one or more recommendations to the marketing team on which countries have potential for growth**

USA is by far the biggest market for Chinook, followed by Canada. However, the average sales per customer is less then most of the other countries. The average sales per customer is the highest in Czech Republic and has a lot of potential to grow.

### 7. Albums vs. Individual Tracks

**Write a query that categorizes each invoice as either an album purchase or not, and calculates the following summary statistics: number of invoices,percentage of invoices**
```{r}
test<-'
WITH
  invoice_first_track AS
    (
    SELECT
      invoice_id,
      MIN(track_id) first_track
    FROM invoice_line
    GROUP BY invoice_id
    ),
  invoiced_tracks AS
    (
    SELECT
      track_id
    FROM invoice_line il
    WHERE il.invoice_id IN (SELECT invoice_id FROM invoice_first_track)
    ),
  album_tracks AS
    (
    SELECT 
      t.track_id
    FROM track t
    WHERE t.album_id= (
                      SELECT 
                        t.album_id
                      FROM track t
                      WHERE t.track_id IN (SELECT first_track FROM invoice_first_track)
                      )
    )
  
SELECT
  album_purchase,
  COUNT(invoice_id) number_of_invoices,
  COUNT(invoice_id)/(SELECT * FROM invoice) percentage_of_invoices
FROM
  (
  SELECT
    ift.*,
    CASE
      WHEN
          (
          SELECT * FROM album_tracks
          EXCEPT
          SELECT * FROM invoiced_tracks
          ) IS NULL
      AND
          (
          SELECT * FROM invoiced_tracks
          EXCEPT
          SELECT * FROM album_tracks
          ) IS NULL
      THEN "yes"
      ELSE "no"
      END AS "album_purchase"
  FROM invoice_first_track ift
  )
GROUP BY 1
'
run_query(test)
```

```{r}
albums_vs_tracks = '
WITH invoice_first_track AS
    (
     SELECT
         il.invoice_id invoice_id,
         MIN(il.track_id) first_track_id
     FROM invoice_line il
     GROUP BY 1
    )
SELECT
    album_purchase,
    COUNT(invoice_id) number_of_invoices,
    CAST(count(invoice_id) AS FLOAT) / (
                                         SELECT COUNT(*) FROM invoice
                                      ) percent
FROM
    (
    SELECT
        ifs.*,
        CASE
            WHEN
                 (
                  SELECT t.track_id FROM track t
                  WHERE t.album_id = (
                                      SELECT t2.album_id FROM track t2
                                      WHERE t2.track_id = ifs.first_track_id
                                     ) 
                  EXCEPT 
                  SELECT il2.track_id FROM invoice_line il2
                  WHERE il2.invoice_id = ifs.invoice_id
                 ) IS NULL
             AND
                 (
                  SELECT il2.track_id FROM invoice_line il2
                  WHERE il2.invoice_id = ifs.invoice_id
                  EXCEPT 
                  SELECT t.track_id FROM track t
                  WHERE t.album_id = (
                                      SELECT t2.album_id FROM track t2
                                      WHERE t2.track_id = ifs.first_track_id
                                     ) 
                 ) IS NULL
             THEN "yes"
             ELSE "no"
         END AS "album_purchase"
     FROM invoice_first_track ifs
    )
GROUP BY album_purchase;
'
run_query(albums_vs_tracks)
```

**Write one to two sentences explaining your findings, and making a prospective recommendation on whether the Chinook store should continue to buy full albums from record companies.**

Album purchases account for almost a quarter of the total sales, so it is inadvisable to change strategy to just purchase the most popular tracks.