---
title: "Guided Project: Designing and Creating a Database"
author: "Chun Hoe Leung"
output: html_notebook
---


### 0. Intro
I will be working with data from *Major League Baseball games* compiled by *Retrosheet*, a non-profit organization that's gathered game statistics going back to the 1800s to today. The data consists of 4 datasets:
-game_log.csv [Download here] (https://dsserver-prod-resources-1.s3.amazonaws.com/376/game_log.csv)
-park_codes.csv [Download here] (https://dsserver-prod-resources-1.s3.amazonaws.com/376/park_codes.csv)
-person_codes.csv [Download here] (https://dsserver-prod-resources-1.s3.amazonaws.com/376/person_codes.csv)
-team_codes.csv [Download here] (https://dsserver-prod-resources-1.s3.amazonaws.com/376/team_codes.csv)
The main file *game_log.csv* has been compiled and pre-cleaned from 127 separate CSV files from Retrosheet. This file has hundreds of data points on each game. The three remaining files are "helper" files and form the basis for three of our normalized tables. 

The goal of this mission is to convert and normalize this data into several separate tables using SQL and create a robust database of game-level statistics.

### 1. Exploring the Data
**Importing Libraries**
```{r}
library(readr)
library(dplyr)
library(stringr)
library(purrr)
library(tidyr)
library(ggplot2)
library(RSQLite)
library(DBI)
```
**Importing Data**
```{r}
# R has trouble guessing some column types, so we explicitly tell it 
# the types of the problem columns

game_log<-read_csv("C:/Users/Familie Leung/Documents/CHL/DataQuest R/Guided Project Designing and Creating a Database/game_log.csv",
                col_types = cols(.default = "c",
                                 v_league = "c", h_league = "c",
                                 `3b_umpire_id` = "c", `3b_umpire_name` = "c",
                                 `2b_umpire_id` = "c", `2b_umpire_name` = "c",
                                 `lf_umpire_id` = "c", `lf_umpire_name` = "c",
                                 `rf_umpire_id` = "c", `rf_umpire_name` = "c",
                                 completion = "c", winning_rbi_batter_id = "c",
                                 winning_rbi_batter_id_name = "c", protest = "c",
                                 v_first_catcher_interference = "c", 
                                 h_first_catcher_interference = "c"))
park_codes<-read_csv("C:/Users/Familie Leung/Documents/CHL/DataQuest R/Guided Project Designing and Creating a Database/park_codes.csv")
person_codes<-read_csv("C:/Users/Familie Leung/Documents/CHL/DataQuest R/Guided Project Designing and Creating a Database/person_codes.csv")
team_codes<-read_csv("C:/Users/Familie Leung/Documents/CHL/DataQuest R/Guided Project Designing and Creating a Database/team_codes.csv")
```
**Exploring the game_log Data**
```{r}
dim(game_log)
head(game_log)
colnames(game_log)
```
**Exploring the park_codes Data**
```{r}
dim(park_codes)
head(park_codes)
colnames(park_codes)
```

**Exploring the person_codes Data**
```{r}
dim(person_codes)
head(person_codes)
colnames(person_codes)
```

**Exploring the team_codes Data**
```{r}
dim(team_codes)
head(team_codes)
colnames(team_codes)
```

**Exploring how the helper files intersects with the main file**

List of ID columns from main file:
```{r}
column_names_game_log<-colnames(game_log)
game_log_id<-column_names_game_log[str_detect(colnames(game_log),"_id")]
game_log_id
```

game_log_id examples:
```{r}
head(game_log[game_log_id])
```

park_codes
```{r}
unique(park_codes$park_id)
```

person_codes
```{r}
example_person<-"boakj901"
person_codes%>%filter(id==example_person)
```

team_codes
```{r}
team_codes[str_detect(colnames(team_codes),"id")]
sum(team_codes$team_id==team_codes$franch_id)
sum(team_codes$team_id==team_codes$franch_id)/nrow(team_codes)
```


**Findings So Far**
1. dimensions
                      records columns id's
dimensions game_log    171907    161    34 (see list above)
dimensions park_codes     252      9     1 (park_id)
dimensions person_codes 20494      7     1 (id)
dimensions team_codes     150      8     2 (team_id,franch_id)

2. game_log only contains FOREIGN KEYS, but no PRIMARY KEY.
3. Both game_log and park_codes contain the park_id column which can serve as a FOREIGN and PRIMARY KEY.
4. person_codes contains personal information regarding the umpire, manager, pitcher and player columns in game_log
5. team_codes has two id columns, none of them is available in game_log file. However, there seems to be an connection between the team_id which consists of 3 letters AND the first 3 letters of the park_id which occurs in game_log and park_codes. Both team_id and franch_id contains a three letter ID. Furthermore, team_id equals 114 out of 150 (76%) to franch_id. 

6. Defensive Positions

In the game log, each player has a defensive position listed, which seems to be a number between 1-10. Doing some research around this, I found this [article] (http://probaseballinsider.com/baseball-instruction/baseball-basics/baseball-basics-positions/) which gives us a list of names for each numbered position:

* Pitcher
* Catcher
* 1st Base
* 2nd Base
* 3rd Base
* Shortstop
* Left Field
* Center Field
* Right Field

The 10th position isn't included, it may be a way of describing a designated hitter that does not field. I can find a retrosheet page that indicates that position 0 is used for this, but we don't have any position 0 in our data. I have chosen to make this an 'Unknown Position' so I'm not including data based on a hunch.

7. Leagues

Wikipedia tells us there are currently two leagues - the American (AL) and National (NL). Upon investigation of the data, we see that there are actually 4 more. After some googling, we come up with:

* NL: National League
* AL: American League
* AA: [American Association](https://en.wikipedia.org/wiki/American_Association_%2819th_century%29)
* FL: [Federal League](https://en.wikipedia.org/wiki/Federal_League)
* PL: [Players League](https://en.wikipedia.org/wiki/Players%27_League)
* UA: [Union Association](https://en.wikipedia.org/wiki/Union_Association)

It also looks like we have about 1000 games where the home team doesn't have a value for league.

### 2. Importing Data into SQLite 

**Use dbWriteTable() to create tables for each of our csv files into a new SQLite database, mlb.db. The table name should be the same as each of the CSV filename without the extension, eg game_log.csv should be imported to a table called game_log.**
```{r}
conn<-dbConnect(SQLite(),"mlb.db")
dbWriteTable(conn=conn,name='game_log',value=game_log,row.names=FALSE,header=TRUE)
dbWriteTable(conn=conn,name='park_codes',value=park_codes,row.names=FALSE,header=TRUE)
dbWriteTable(conn=conn,name='person_codes',value=person_codes,row.names=FALSE,header=TRUE)
dbWriteTable(conn=conn,name='team_codes',value=team_codes,row.names=FALSE,header=TRUE)
```
**Check > Listing tables in the database**
```{r}
dbListTables(conn)
```
**Create a new column in the game_log table called game_id**
```{r}
SQLite_shell_query1<-"ALTER TABLE game_log ADD COLUMN game_id TEXT;"
dbExecute(conn,SQLite_shell_query1)
```
**Use SQL string concatenation to update the new columns with a unique ID using the Retrosheet format outlined above.**
```{r}
SQLite_shell_query2<-"
  UPDATE game_log 
  SET game_id=h_name||date||number_of_game;
  /* WHERE prevents this if it has already been done */
  /*WHERE game_id IS NULL;*/
"
dbExecute(conn,SQLite_shell_query2)
```
**Check game_id**
```{r}
RSQLite_library_query1<-'
SELECT
  game_id,
  date,
  h_name,
  number_of_game
FROM game_log
LIMIT 10
'
check1<-dbGetQuery(conn,RSQLite_library_query1)
check1
```

### 3. Looking for Normalization Opportunities

**Looking at the various files, look for opportunities to normalize the data and record your observations in a markdown cell**
The following are opportunities for normalization of our data:

* In person_codes, all the debut dates will be able to be reproduced using game log data.
* In team_codes, the start, end and sequence columns will be able to be reproduced using game log data.
* In park_codes, the start and end years will be able to be reproduced using game log data. While technically the state is an attribute of the city, we might not want to have a an incomplete city/state table so we will leave this in.
* There are lots of places in game log where we have a player ID followed by the players name. We will be able to remove this and use the name data in person_codes
* In game_log, all offensive and defensive stats are repeated for the home team and the visiting team. We could break these out and have a table that lists each game twice, one for each team, and cut out this column repetition.
* Similarly, in game_log, we have a listing for 9 players on each team with their positions - we can remove these and have one table that tracks player appearances and their positions.
* We can do a similar thing with the umpires from game_log, instead of listing all four positions as columns, we can put the umpires either in their own table or make one table for players, umpires and managers.
* We have several awards in game_log like winning pitcher and losing pitcher. We can either break these out into their own table, have a table for awards, or combine the awards in with general appearances like the players and umpires.

### 4. Planning a Normalization Schema
Done

### 5. Creating Tables without Foreign Key Relations

**Create the person table with columns and primary key as shown in the schema diagram**
```{r}
SQLite_shell_query3<-'
CREATE TABLE IF NOT EXISTS person (
person_id TEXT,
first_name TEXT,
last_name TEXT,
PRIMARY KEY (person_id)
);
'
dbExecute(conn,SQLite_shell_query3)
```
**Insert the data in person table from the person_codes table**
```{r}
SQLite_shell_query4<-'
INSERT OR IGNORE INTO person 
SELECT
  id,
  first,
  last
FROM person_codes;
'
dbExecute(conn,SQLite_shell_query4)
```
**Check person table**
```{r}
RSQLite_library_query2<-'
SELECT * FROM person'
check2<-dbGetQuery(conn,RSQLite_library_query2)
check2
```
**Create the park table with columns and primary key as shown in the schema diagram**
```{r}
SQLite_shell_query5<-'
CREATE TABLE IF NOT EXISTS park (
park_id TEXT PRIMARY KEY,
name TEXT,
nickname TEXT,
city TEXT,
state TEXT,
notes TEXT
);
'
dbExecute(conn,SQLite_shell_query5)
```
**Insert the data in park table from the park_codes table**
```{r}
SQLite_shell_query6<-'
INSERT OR IGNORE INTO park 
SELECT
  park_id,
  name,
  aka,
  city,
  state,
  notes
FROM park_codes;
'
dbExecute(conn,SQLite_shell_query6)
```
**Check park table**
```{r}
RSQLite_library_query3<-'
SELECT * FROM park'
check3<-dbGetQuery(conn,RSQLite_library_query3)
check3
```
**Create the league table with columns and primary key as shown in the schema diagram**
```{r}
SQLite_shell_query7<-'
CREATE TABLE IF NOT EXISTS league (
league_id TEXT PRIMARY KEY,
name TEXT
);
'
dbExecute(conn,SQLite_shell_query7)
```
**Insert the data in park table from the park_codes table**
```{r}
SQLite_shell_query8<-'
INSERT OR IGNORE INTO league 
VALUES
("NL","National League"),
("AL","American League"),
("AA","American Association"),
("FL","Federal League"),
("PL","Players League"),
("UA","Union Association");
'
dbExecute(conn,SQLite_shell_query8)
```

**Check park table**
```{r}
RSQLite_library_query4<-'
SELECT * FROM league'
check4<-dbGetQuery(conn,RSQLite_library_query4)
check4
```
**Create the appearance_type table with columns and primary key as shown in the schema diagram**
Can't create this table because I don't have access to appearance_type.csv.

### 6. Adding the Team and Game Tables
**Create the team table with columns, primary key, and foreign key as shown in the schema diagram.**
```{r}
SQLite_shell_query9<-'
CREATE TABLE IF NOT EXISTS team (
team_id TEXT PRIMARY KEY,
league_id TEXT,
city TEXT,
nickname TEXT,
franch_id TEXT,
FOREIGN KEY (league_id) REFERENCES league(league_id)
);
'
dbExecute(conn,SQLite_shell_query9)
```
**Insert the data in team table from the team_codes table**
```{r}
SQLite_shell_query10<-'
INSERT OR IGNORE INTO team 
SELECT
  team_id,
  league,
  city,
  nickname,
  franch_id
FROM team_codes;
'
dbExecute(conn,SQLite_shell_query10)
```
**Check team table**
```{r}
RSQLite_library_query5<-'
SELECT * FROM team'
check5<-dbGetQuery(conn,RSQLite_library_query5)
check5
``` 
**Create the game table with columns, primary key, and foreign key as shown in the schema diagram.**
```{r}
SQLite_shell_query11<-'
CREATE TABLE IF NOT EXISTS game (
  game_id TEXT PRIMARY KEY,
  date TEXT,
  number_of_game INTEGER,
  park_id TEXT,
  length_outs INTEGER,
  day BOOLEAN,
  completion TEXT,
  forefeit TEXT,
  protest TEXT,
  attendance INTEGER,
  length_minutes INTEGER,
  additional_info TEXT,
  acquisition_info TEXT,
  FOREIGN KEY (park_id) REFERENCES park(park_id)
  );
'
dbExecute(conn,SQLite_shell_query11)
```
**Insert the data in game table from the game_log table**
```{r}
SQLite_shell_query12<-'
INSERT OR IGNORE INTO game 
SELECT
  game_id,
  date,
  number_of_game,
  park_id,
  length_outs,
  CASE
      WHEN day_night = "D" THEN 1
      WHEN day_night = "N" THEN 0
      ELSE NULL
      END
      AS day,
  completion,
  forefeit,
  protest,
  attendance,
  length_minutes,
  additional_info,
  acquisition_info
FROM game_log;
'
dbExecute(conn,SQLite_shell_query12)
```
**Check game table**
```{r}
RSQLite_library_query6<-'
SELECT * FROM game'
check6<-dbGetQuery(conn,RSQLite_library_query6)
check6
```

### 7. Adding the Team Appearance Tables
**Create the team_appearance table with columns, primary key, and foreign keys as shown in the schema diagram.**
```{r}
SQLite_shell_query13<-'
CREATE TABLE IF NOT EXISTS team_appearance (
    team_id TEXT,
    game_id TEXT,
    home BOOLEAN,
    league_id TEXT,
    score INTEGER,
    line_score TEXT,
    at_bats INTEGER,
    hits INTEGER,
    doubles INTEGER,
    triples INTEGER,
    homeruns INTEGER,
    rbi INTEGER,
    sacrifice_hits INTEGER,
    sacrifice_flies INTEGER,
    hit_by_pitch INTEGER,
    walks INTEGER,
    intentional_walks INTEGER,
    strikeouts INTEGER,
    stolen_bases INTEGER,
    caught_stealing INTEGER,
    grounded_into_double INTEGER,
    first_catcher_interference INTEGER,
    left_on_base INTEGER,
    pitchers_used INTEGER,
    individual_earned_runs INTEGER,
    team_earned_runs INTEGER,
    wild_pitches INTEGER,
    balks INTEGER,
    putouts INTEGER,
    assists INTEGER,
    errors INTEGER,
    passed_balls INTEGER,
    double_plays INTEGER,
    triple_plays INTEGER,
    PRIMARY KEY (team_id, game_id),
    FOREIGN KEY (team_id) REFERENCES team(team_id),
    FOREIGN KEY (game_id) REFERENCES game(game_id),
    FOREIGN KEY (team_id) REFERENCES team(team_id)
  );
'
dbExecute(conn,SQLite_shell_query13)
```
**Insert the data in team_appearance table from the game_log table**
```{r}
SQLite_shell_query14<-'
INSERT OR IGNORE INTO team_appearance
    SELECT
        h_name,
        game_id,
        1 AS home,
        h_league,
        h_score,
        h_line_score,
        h_at_bats,
        h_hits,
        h_doubles,
        h_triples,
        h_homeruns,
        h_rbi,
        h_sacrifice_hits,
        h_sacrifice_flies,
        h_hit_by_pitch,
        h_walks,
        h_intentional_walks,
        h_strikeouts,
        h_stolen_bases,
        h_caught_stealing,
        h_grounded_into_double,
        h_first_catcher_interference,
        h_left_on_base,
        h_pitchers_used,
        h_individual_earned_runs,
        h_team_earned_runs,
        h_wild_pitches,
        h_balks,
        h_putouts,
        h_assists,
        h_errors,
        h_passed_balls,
        h_double_plays,
        h_triple_plays
    FROM game_log

UNION

    SELECT    
        v_name,
        game_id,
        0 AS home,
        v_league,
        v_score,
        v_line_score,
        v_at_bats,
        v_hits,
        v_doubles,
        v_triples,
        v_homeruns,
        v_rbi,
        v_sacrifice_hits,
        v_sacrifice_flies,
        v_hit_by_pitch,
        v_walks,
        v_intentional_walks,
        v_strikeouts,
        v_stolen_bases,
        v_caught_stealing,
        v_grounded_into_double,
        v_first_catcher_interference,
        v_left_on_base,
        v_pitchers_used,
        v_individual_earned_runs,
        v_team_earned_runs,
        v_wild_pitches,
        v_balks,
        v_putouts,
        v_assists,
        v_errors,
        v_passed_balls,
        v_double_plays,
        v_triple_plays
    from game_log;
'
dbExecute(conn,SQLite_shell_query14)
```
**Check team_appearance table**
```{r}
RSQLite_library_query7<-'
SELECT * FROM team_appearance'
check7<-dbGetQuery(conn,RSQLite_library_query7)
check7
```
**Another check**
```{r}
check_query_8 <- "
  SELECT * FROM team_appearance
  WHERE game_id = (
                   SELECT MIN(game_id) from game
                  )
     OR game_id = (
                   SELECT MAX(game_id) from game
                  )
  ORDER By game_id, home;
"
check <- dbGetQuery(conn, check_query_8)
head(check)
```

### 8. Adding the Person Appearance Tables

**Create the person_appearance table with columns, primary key, and foreign keys as shown in the schema diagram.**
```{r}
drop_person_appearance_precaution <- "DROP TABLE IF EXISTS person_appearance"
dbExecute(conn, drop_person_appearance_precaution)

SQLite_shell_query15<-'
  CREATE TABLE person_appearance (
      appearance_id INTEGER PRIMARY KEY,
      person_id TEXT,
      team_id TEXT,
      game_id TEXT,
      appearance_type_id,
      FOREIGN KEY (person_id) REFERENCES person(person_id),
      FOREIGN KEY (team_id) REFERENCES team(team_id),
      FOREIGN KEY (game_id) REFERENCES game(game_id),
      FOREIGN KEY (appearance_type_id) REFERENCES appearance_type(appearance_type_id)
  );
'
dbExecute(conn,SQLite_shell_query15)
```
**Insert the data in person_appearance table from the game_log table**
```{r}
SQLite_shell_query16<-'
  INSERT OR IGNORE INTO person_appearance (
      game_id,
      team_id,
      person_id,
      appearance_type_id
  ) 
      SELECT
          game_id,
          NULL,
          hp_umpire_id,
          "UHP"
      FROM game_log
      WHERE hp_umpire_id IS NOT NULL    
  
  UNION
  
      SELECT
          game_id,
          NULL,
          [1b_umpire_id],
          "U1B"
      FROM game_log
      WHERE "1b_umpire_id" IS NOT NULL
  
  UNION
  
      SELECT
          game_id,
          NULL,
          [2b_umpire_id],
          "U2B"
      FROM game_log
      WHERE [2b_umpire_id] IS NOT NULL
  
  UNION
  
      SELECT
          game_id,
          NULL,
          [3b_umpire_id],
          "U3B"
      FROM game_log
      WHERE [3b_umpire_id] IS NOT NULL
  
  UNION
  
      SELECT
          game_id,
          NULL,
          lf_umpire_id,
          "ULF"
      FROM game_log
      WHERE lf_umpire_id IS NOT NULL
  
  UNION
  
      SELECT
          game_id,
          NULL,
          rf_umpire_id,
          "URF"
      FROM game_log
      WHERE rf_umpire_id IS NOT NULL
  
  UNION
  
      SELECT
          game_id,
          v_name,
          v_manager_id,
          "MM"
      FROM game_log
      WHERE v_manager_id IS NOT NULL
  
  UNION
  
      SELECT
          game_id,
          h_name,
          h_manager_id,
          "MM"
      FROM game_log
      WHERE h_manager_id IS NOT NULL
  
  UNION
  
      SELECT
          game_id,
          CASE
              WHEN h_score > v_score THEN h_name
              ELSE v_name
              END,
          winning_pitcher_id,
          "AWP"
      FROM game_log
      WHERE winning_pitcher_id IS NOT NULL
  
  UNION
  
      SELECT
          game_id,
          CASE
              WHEN h_score < v_score THEN h_name
              ELSE v_name
              END,
          losing_pitcher_id,
          "ALP"
      FROM game_log
      WHERE losing_pitcher_id IS NOT NULL
  
  UNION
  
      SELECT
          game_id,
          CASE
              WHEN h_score > v_score THEN h_name
              ELSE v_name
              END,
          saving_pitcher_id,
          "ASP"
      FROM game_log
      WHERE saving_pitcher_id IS NOT NULL
  
  UNION
  
      SELECT
          game_id,
          CASE
              WHEN h_score > v_score THEN h_name
              ELSE v_name
              END,
          winning_rbi_batter_id,
          "AWB"
      FROM game_log
      WHERE winning_rbi_batter_id IS NOT NULL
  
  UNION
  
      SELECT
          game_id,
          v_name,
          v_starting_pitcher_id,
          "PSP"
      FROM game_log
      WHERE v_starting_pitcher_id IS NOT NULL
  
  UNION
  
      SELECT
          game_id,
          h_name,
          h_starting_pitcher_id,
          "PSP"
      FROM game_log
      WHERE h_starting_pitcher_id IS NOT NULL;
'
dbExecute(conn,SQLite_shell_query16)
```
**permutations**
```{r}
for (letter in c("h", "v")) {
  for (num in 1:9) {
    template <- '
      INSERT INTO person_appearance (
          game_id,
          team_id,
          person_id,
          appearance_type_id
      ) 
          SELECT
              game_id,
              %s_name,
              %s_player_%f_id,
              "O%f"
          FROM game_log
          WHERE %s_player_%f_id IS NOT NULL
      
      UNION
      
          SELECT
              game_id,
              %s_name,
              %s_player_%f_id,
              "D" || CAST(%s_player_%f_def_pos AS INT)
          FROM game_log
          WHERE %s_player_%f_id IS NOT NULL;
    '
    # replace all of the %s and %f with the correct letter number
    template <- gsub("%s", letter, template, fixed = TRUE)
    template <- gsub("%f", num, template, fixed = TRUE)
    
    dbExecute(conn, template)
  }
}
```

This particular solution incorporates a bit of regular expression and some nested for loops. If you need to review for-loops again, refer to our older mission on [control structures](https://www.dataquest.io/m/338-working-with-control-structures/). Regular expression is a technique where we look for specific patterns in a character vector and replace it with another pattern. In this case, we need to replace all the instances of "%s" and "%f" with a new letter and number, respectively. In particular, the `gsub()` function replaces *all* instances of "%s" and "%f" respectively. With each run of the nested loop, we get a properly formatted query that fulfills our needs. Using loops is one potential solution, but it's also possible to vectorize this work and recreate what we need.


### 9. Removing the Original Tables
```{r}
# Check the current status of the db file
dbListTables(conn)

```

```{r}
# Iterate through each of the tables
tables <- c("game_log", "park_codes",
            "team_codes", "person_codes")
for (t in tables) {
  drop_command = sprintf("DROP TABLE %s", t)
  dbExecute(conn, drop_command)
}
```

```{r}
# Make sure that everything is gone
dbListTables(conn)
```
**Disconnect**
```{r}
dbDisconnect(conn)
```

