---
title: "Guided Project: Analyzing CIA Factbook Data Using SQLite and R"
author: "Chun Hoe Leung"
output: html_notebook
---

### 1. Intro
For this project, I'll work with data from the [CIA World Factbook] (https://www.cia.gov/library/publications/the-world-factbook/), a compendium of statistics about all of the countries on Earth. The Factbook contains demographic information like:

1. population - The population as of 2015.
2. population_growth - The annual population growth rate, as a percentage.
3. area - The total land and water area.

The data are publicly available and can be accessed [here](https://github.com/factbook/factbook.sql/releases). The file I'll be working with is *factbook.db*. I'll practice some newfound R and SQLite skills on this database and create some useful visualizations of the data.

### 2. Overview of the Data
**Import libraries RSQLite and DBI**
```{r}
library(RSQLite)
library(DBI)
```
**Import factbook.db and List all tables**
```{r}
conn<-dbConnect(SQLite(),"C:/Users/Familie Leung/Documents/CHL/DataQuest R/Guided Project Analyzing CIA Factbook Data Using SQLite and R/factbook.db")
dbListTables(conn)
```
**Explore each column in table facts**
```{r}
query0<-"PRAGMA TABLE_INFO(facts)"
result0<-dbGetQuery(conn,query0)
print(result0)
```
**Explore number of rows in table facts**
```{r}
query00<-"SELECT COUNT(id) FROM facts"
result00<-dbGetQuery(conn,query00)
print(result00)
```


**Write and run a query that returns the first 5 rows of the facts table in the database.**
```{r}
query1<-"SELECT * FROM facts LIMIT 5"
result1<-dbGetQuery(conn,query1)
print(result1)
```


### 3. Summary Statistics
The descriptions for some of the columns:
1. name > The name of the country.
2. area > The total land and sea area of the country.
3. population > The country's population.
4. population_growth > The country's population growth as a percentage.
5. birth_rate > The country's birth rate, or the number of births a year per 1,000 people.
6. death_rate > The country's death rate, or the number of death a year per 1,000 people.
7. area > The country's total area (both land and water).
8. area_land > The country's land area in square kilometers (https://www.cia.gov/library/publications/the-world-factbook/rankorder/2147rank.html).
9. area_water > The country's waterarea in square kilometers.

**Calculate summary statistics MIN() and MAX() for the 'population' and 'population_growth' variables and check for any outlier countries.**
```{r}
query2<-"SELECT MIN(population),MAX(population),MIN(population_growth),MAX(population_growth) FROM facts"
result2<-dbGetQuery(conn,query2)
result2
```

### 4. Exploring Outliers
**Write a query that returns the countrie(s) with a population of 0**
```{r}
query3<-"SELECT * FROM facts WHERE (population=0)"
result3<-dbGetQuery(conn,query3)
result3
```
**Write a query that returns the countrie(s) with a population of 7256490011**
```{r}
query4<-"SELECT * FROM facts WHERE population=7256490011"
result4<-dbGetQuery(conn,query4)
result4
```
**Observations so far**
The population variable contains outliers varying from 0 to 7.2billion.
-Antartica has a population which equals 0
-World has a population which equals 7.2billion

### 5. Histograms
**Exclude the countries Antartica and World in a revised query**
```{r}
query5<-"SELECT population,population_growth,birth_rate,death_rate FROM facts WHERE name!='Antarctica' AND name!='World'"
facts<-dbGetQuery(conn,query5)
facts
```
**Import libraries tidyr and ggplot2**
```{r}
library(tidyr)
library(ggplot2)
```
**Tidy data**
```{r}
tidy_facts<-facts%>%
  gather(.,key="variable",value="value")
tidy_facts
```

**Generate histograms for columns: population, population_growth, birth_rate,death_rate**
```{r}
ggplot(data=tidy_facts)+aes(x=value,fill=variable)+geom_histogram()+facet_grid(~variable)+labs(title="Histogram")
```
Histogram birth_rate
```{r}
ggplot(data=facts)+aes(x=birth_rate)+geom_histogram()+theme(panel.background = element_rect(fill="white"))+labs(title="Histogram birth_rate",x="number of births a year per 1,000 people")
```
Histogram death_rate
```{r}
ggplot(data=facts)+aes(x=death_rate)+geom_histogram()+labs(title="Histogram deathth_rate",x="number of death a year per 1,000 people")
```
Histogram population
```{r}
ggplot(data=facts)+aes(x=population)+geom_histogram(bins=20)+labs(title="Histogram population")
```
Histogram population_growth
```{r}
ggplot(data=facts)+aes(x=population_growth)+geom_histogram()+labs(title="Histogram population_growth")
```

